Simulating High-Dimensional Multivariate Data using the \pkg{bigsimr} Package

### 2021-06-03 000-introduction.Rmd
get bookdown cross refs working
Final push before pre-JSS submit internal review

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness --- often to compare to methods. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19]. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with flexible, robust software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have largely remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions for discrete marginal distributions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation is required.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution. For our method, we developed a high-performance implementation: the `bigsimr` R package, written in native Julia code.

Our study proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. We proceed by providing an illustrative LD `bigsimr` workflow. Next we conduct MC studies under various bivariate distributional assumptions to evaluate performance and accuracy. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.

>>>>

\setstretch{2.0}

Massive high-dimensional (HD) data sets are now commonplace in many areas of scientific inquiry. As new methods are developed for data analysis, a fundamental challenge lies in designing and conducting simulation studies to assess the operating characteristics of proposed methodology --- such as false positive rates, statistical power, interval coverage, and robustness. Further, efficient simulation empowers statistical computing strategies, such as the parametric bootstrap [@Chernick2008] to simulate from a hypothesized null model, providing inference in analytically challenging settings. Such Monte Carlo (MC) techniques become difficult for HD dependent data using existing algorithms and tools. This is particularly true when simulating massive multivariate, non-normal distributions, arising in many fields of study.

As many have noted, it can be vexing to simulate dependent, non-normal/discrete data, even for low-dimensional (LD) settings [@MB13; @XZ19]. For continuous non-normal LD multivariate data, the well-known NORmal To Anything (NORTA) algorithm [@Cario1997] and other copula approaches [@Nelsen2007] are well-studied, with existing software available [@Yan2007; @Chen2001]. Yet these approaches do not scale in a timely fashion to HD problems [@Li2019gpu]. For discrete data, early simulation strategies had major flaws, such as failing to obtain the full range of possible correlations (e.g., admitting only positive correlations: see @Park1996). While more recent approaches [@MB13; @Xia17; @BF17] have remedied this issue for LD problems, the existing tools are not designed to scale to high dimensions.

Another central issue lies in characterizing dependence between components in the HD random vector. The choice of correlation in practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g., non-normal, discrete, infinite support, etc). For normal data, the Pearson product-moment correlation describes the dependency perfectly. As we will see, however, simulating arbitrary random vectors that match a target Pearson correlation matrix is computationally intense [@Chen2001; @Xia17]. On the other hand, an analyst might consider use of nonparametric correlation measures to better characterize monotone, non-linear dependence, such as Spearman's $\rho$ and Kendall's $\tau$. Throughout, we focus on matching these nonparametric dependence measures, as our aim lies in modeling non-normal data and these rank-based measures possess invariance properties favorable in our proposed methodology. We do, however, implement Pearson matching, but several layers of approximation are required.

With all this in mind, we present a scalable, flexible multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula in the spirit of the NORTA procedure. As we will describe in more detail, the algorithm's design leverages useful properties of nonparametric correlation measures, namely invariance under monotone transformation and well-known closed-form relationships between dependence measures for the multivariate normal (MVN) distribution. For our method, we developed a high-performance implementation: the `Bigsimr` Julia package, with R and Python interfaces `bigsimr`.

This article proceeds by providing background information, including a description of a motivating example application: RNA-sequencing (RNA-seq) breast cancer data. Then we describe and justify our simulation methodology and related algorithms. We proceed by providing an illustrative LD `bigsimr` workflow. Next we conduct MC studies under various bivariate distributional assumptions to evaluate performance and accuracy. After the MC evaluations, we simulate random vectors motivated by our RNA-seq example, evaluate the accuracy, and provide example statistical computing tasks, namely MC estimation of joint probabilities and evaluating HD correlation estimation efficiency. Finally, we discuss the method's utility, limitations, and future directions.

### 2021-06-03 010-background.Rmd
Final push before pre-JSS submit internal review

The `bigsimr` `R` package presented here provides multiple algorithms that operate with HD multivariate data; however, all these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i, i=1,\ldots,d$.

When designing this methodology, we developed the following properties to guide our effort. We divide the properties into two categories: (1) basic properties (BP) and "scalability" properties (SP). The BPs are adapted from an existing criteria due to @Nik13a. A suitable simulation strategy should possess the following properties:

@@

* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
* BP2: Flexible dependence, meaning that the number of bivariate marginals can be equal to the number of dependence parameters.
* BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete margins.

@@

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby capturing nonlinear yet monotone relationships. Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$ (concordance meaning that orderings have the same direction, e.g., if $X_i < X_j$, then $Y_i < Y_j$). Note that concordance is determined by the ranks of the values, not the values themselves.

Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates. As we will describe more fully in the [Algorithms](#algorithms) section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).

@@

There is a connection of our simulation strategy to Gaussian *copulas* [@Nelsen2007]. A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability. Consequently, the study of copulas is an important and active area of statistical theory and practice.

For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), 
\label{eq:copula}
\end{equation}

$x_i \in {\mathbb R}, i=1,\ldots,d.$

A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, where $\Phi(\cdot)$ is the standard normal CDF. This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix ${\bf R_P}$. Since the marginals are standardized to have unit variance, however, ${\bf R_P}$ is a Pearson correlation matrix. If $F_{{\bf R}}$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_{{\bf R}}$ is defined through

\begin{equation}
\label{eq:gauss}
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

Here the copula function $C_{{\bf R}}$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz. $Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R_P})$.

## Other Multivariate Simulation R Packages

To place `bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other `R` multivariate simulation packages.  The `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This familiar procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. Recently, the high-performance package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD packages exist, such as `GenOrd`. For copula computation, the full-featured package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table \@ref(tab:compare-table) summarizes the above discussion.

\begin{table}[]
\centering
\caption{Selected feature comparison of bigsimr to existing R packages for multivariate simulation.}
\label{compare-table}
\begin{tabular}{@{}llll@{}}
R Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

>>>>

The `Bigsimr` package presented here provides multiple high-performance algorithms that operate with HD multivariate data. All these algorithms were originally designed to support a single task: to generate random vectors drawn from multivariate probability distributions with given marginal distributions and dependency metrics. Specifically, our goal is to efficiently simulate a large number, $B$, of HD random vectors ${\bf Y}=(Y_1, \ldots, Y_d)^\top$ with *correlated* components and heterogeneous marginal distributions, described via cumulative distribution functions (CDFs) $F_i, i=1,\ldots,d$.

When designing this methodology, we developed the following properties to guide our effort. We divide the properties into two categories: (1) basic properties (BP) and 'scalability' properties (SP). The BPs are adapted from an existing criteria due to @Nik13a. A suitable simulation strategy should possess the following properties:

@@

* BP1: A wide range of dependences, allowing both positive and negative values, and, ideally, admitting the full range of possible values.
* BP2: Flexible dependence, meaning that the number of dependence parameters can be equal to the number of bivariate pairs in the vector.
* BP3: Flexible marginal modeling, generating heterogeneous data --- including mixed continuous and discrete marginal distributions.

@@

Two nonparametric, rank-based measures common in practice are Spearman's correlation (denoted $\rho_S$) and Kendall's $\tau$. Spearman's $\rho_S$ has an appealing correspondence as the Pearson correlation coefficient on *ranks* of the values, thereby capturing nonlinear yet monotone relationships. Kendall's $\tau$, on the other hand, is the difference in probabilities of concordant and discordant pairs of observations, $(X_i, Y_i)$ and $(X_j, Y_j)$ (concordance meaning that orderings have the same direction, e.g., if $X_i < X_j$, then $Y_i < Y_j$). Note that concordance is determined by the ranks of the values, not the values themselves. Both $\tau$ and $\rho_S$ are *invariant under monotone transformations* of the underlying random variates. As we will describe more fully in the [Algorithms](#algorithms) section, this property enables matching rank-based correlations with speed (SP1) and accuracy (SP2).

@@

There is a connection of our simulation strategy to Gaussian *copulas* [@Nelsen2007]. A copula is a distribution function on $[0,1]^d$ that describes a multivariate probability distribution with standard uniform marginals. This provides a powerful, natural way to characterize joint probability. Consequently, the study of copulas is an important and active area of statistical theory and practice. For any random vector ${\bf X}=(X_1, \ldots, X_d)^\top$ with CDF $F$ and marginal CDFs $F_i$ there is a copula function $C(u_1, \ldots, u_d)$ satisfying

\begin{equation}
F(x_1, \ldots, x_d) = {\mathbb P}(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \quad x_i \in {\mathbb R}, i=1,\ldots,d.
\label{eq:copula}
\end{equation}

A Gaussian copula has marginal CDFs that are all standard normal, $F_i = \Phi, \forall \, i$, where $\Phi(\cdot)$ is the standard normal CDF. This representation corresponds to a multivariate normal (MVN) distribution with standard normal marginal distributions and covariance matrix $R_P$. Since the marginals are standardized to have unit variance, however, $R_P$ is a Pearson correlation matrix. If $F_R$ is the CDF of such a multivariate normal distribution, then the corresponding Gaussian copula $C_R$ is defined through

\begin{equation}
\label{eq:gauss}
F_R(x_1, \ldots, x_d) = C_R(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

Here the copula function $C_R$ is the familiar multivariate normal CDF of the random vector $(\Phi(X_1), \ldots, \Phi(X_d))$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.

Sklar's Theorem [@Sklar1959; @Ubeda-Flores2017] guarantees that given inverse CDFs $F_i^{-1}$s and a valid correlation matrix (within the Frechet bounds) a random vector can be obtained via transformations involving copula functions. For example, using Gaussian copulas, we can construct a random vector ${\bf Y}  = (Y_1, \ldots,  Y_d)^\top$ with $Y_i \sim F_i$, viz. $Y_i = F_i^{-1}(\Phi(X_i)), i=1, \ldots, d$, where $(X_1, \ldots, X_d) \sim N_d({\bf 0}, R_P)$.

## Other Multivariate Simulation Packages

To place `Bigsimr`'s implementation and algorithms in context with existing approaches, here we qualitatively compare to other multivariate simulation packages.  In `R`, the `MASS` packages `mvnorm()` has long been available to produce random vectors from multivariate normal (and $t$) distributions. This procedure cannot simulate from arbitrary margins and dependency measures, as well as fails to readily scale to high dimensions. SAS PROC SIMNORMAL and Python's NumPy `random.multivariate_normal` both do similar tasks. Recently, the high-performance `R` package `mvnfast` [@Fasiolo2016] has been released that provides HD multivariate normal generation, but this lacks flexibility in marginal models and dependency modeling. For dependent discrete data, LD `R` packages exist, such as `GenOrd`. For copula computation, the full-featured `R` package `copula` provides many different copulas and a fairly general LD random generator `rCopula()`. High-dimesional random vectors via `copula` takes an impractical amount of computing time [@Li2019gpu]. The `nortaRA` R package matches Pearson coefficients near exactly and is reasonable for LD settings. Finally, we know of no other `R` package that can produce HD random vectors with flexible margins and dependence modeling. Table \@ref(tab:compare-table) summarizes the above discussion.

\begin{table}[h]
\centering
\caption{\label{tab:compare-table}Comparison of selected features of \pkg{Bigsimr} to existing packages for multivariate simulation.}
\begin{tabular}{@{}llll@{}}
Package & Scales to HD & Hetereogenous Margins & Flexible Dependence \\ \midrule
MASS      & No           & No                    & No                  \\
mvnfast   & Yes          & No                    & No                  \\
GenOrd    & No           & Discrete margins      & Yes                 \\
copula    & No           & Yes                   & No                  \\
nortaRA   & No           & Yes                   & No                  \\
bigsimr   & Yes          & Yes                   & Yes                
\end{tabular}
\end{table}

\clearpage

### 2021-06-03 020-simulation-algorithm.Rmd
Final push before pre-JSS submit internal review

where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables. For HD simulation, exact evaluation may become too costly to enable practical simulation studies. Our remedy to enable HD Pearson matching in `bigsimr` is Hermite polynomial approximation of these $\choose{d}{2}$ double integrals, as suggested by @XZ19.

@@

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. Yet the choice of copula-based simulation algorithm affects which joint distributions may be simulated. Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75]. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@GH02 conducted a Monte Carlo study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides guidance that our nearest positive definite (PD) augmented approach has promise provide adequate accuracy if our input matching scheme returns an indefinite matrix.

## bigsimr rvec: Random vector generator {#rand-vec-gen}

@@

There is an increasing probability of encountering a inadmissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix. In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

*Step 3: Gaussian copula.* The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins. Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation. Step 3 requires a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals. The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\Phi(\cdot)$. Finally, the quantile functions $F^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.

>>>>

where $\phi(\cdot)$ is the joint probability distribution function of two correlated standard normal variables. For HD simulation, exact evaluation may become too costly to enable practical simulation studies. Our remedy to enable HD Pearson matching in `bigsimr` is Hermite polynomial approximation of these $\binom{d}{2}$ double integrals, as suggested by @XZ19.

@@

Sklar's theorem provides a useful characterization of multivariate distributions through copulas. In practice, however, the particulars of the simulation algorithm affect which joint distributions may be simulated. Even in LD spaces (e.g., $d=3$), there exist valid multivariate distributions with *feasible* Pearson correlation matrices that NORTA cannot match exactly [@LH75]. This occurs when the bivariate transformations are applied to find the input correlation matrix, yet when combined the resultant matrix is indefinite. These situations do occur, even using exact analytic calculations. Such problematic target correlation matrices are termed *NORTA defective*.

@GH02 conducted a study to estimate the probability of encountering NORTA defective matrices while increasing the dimension $d$. They found that for what is now considered low-to-moderate dimensions ($d \approx 20$), almost *all* feasible matrices are NORTA defective. This stems from the concentration of measure near the boundary of the space of all possible correlation matrices as dimension increases. Unfortunately, it is precisely near this boundary that NORTA defective matrices reside.

There is hope, however, as @GH02 also showed that replacing an indefinite input correlation matrix with a close proxy will give approximate matching to the target --- with adequate performance for moderate $d$. This provides insight that our nearest positive definite (PD) augmented approach has promise to provide adequate accuracy if our input matching scheme returns an indefinite Pearson correlation matrix.

## Random vector generator {#rand-vec-gen}

@@

There is an increasing probability of encountering an inadmissible correlation matrix as dimension increases. In our experience, the mapping step for large $d$ almost always produces a $R_X$ that is not a correlation matrix. In Section \@ref(package) we provide a basic description of how to quantify and control the approximation error. Further, the RNA-seq example in Section \@ref(examples) provides an illustration of this in practice.

*Step 3: Gaussian copula.* The final step implements a NORTA-inspired, Gaussian copula approach to produce the desired margins. Steps 1 and 2 determine the MVN Pearson correlation values that will eventually match the target correlation. Step 3 requires a fast MVN simulator, a standard normal CDF, and well-defined quantile functions for marginals. The MVN is transformed to a copula (distribution with standard uniform margins) by applying the normal CDF $\Phi(\cdot)$. Finally, the quantile functions $F_i^{-1}$ are applied across the margins to return the desired random vector ${\bf Y}$.

### 2021-06-03 030-bigsimr-package.Rmd
Final push before pre-JSS submit internal review

# The \pkg{bigsimr} R package {short-title="bigsimr-pkg" #package}

@@

This section describes a LD random vector simulation workflow via the `bigsimr` `R` package. Our `R` package provides an interface to the native code written in Julia (registered as the `Bigsimr` Julia package. In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` that interfaces with the Julia `Bigsimr` package. The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).

The subsections below describe the basic use of `bigsimr` by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983]. This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.

@@

Below, we describe a minimal working example --- a bivariate simulation of two airquality variables: `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.

```{r ch030-air-quality-data, echo=TRUE}
df <- airquality %>% select(Temp, Ozone) %>% drop_na()
```

```{r ch030-aq-glimpse}
glimpse(df)
```

@@

As mentioned, the user must decide how to describe correlation based on the particulars of the problem. For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$. We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for the distribution at hand (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance). Note that hese estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.

@@

Once a target correlation matrix is specified, first convert the dependency type to Pearson to correctly specify the MVN inputs. Then, check that the converted matrix $R_X$ is a valid correlation matrix (PD and 1s on the diagonal). For this bivariate example, $R_X$ is a valid correlation matrix. Typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).

@@

A single `rvec` execution simulates the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`. Note that the target input in specified using the pre-computed MVN correlation matrix.

@@

Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

@@

>>>>

# The \pkg{Bigsimr} package {short-title="Bigsimr-pkg" #package}

@@

This section describes a bivariate random vector simulation workflow via the `bigsimr` `R` package. This `R` package provides an interface to the native code written in Julia (registered as the `Bigsimr` Julia package). In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a Python interface `bigsimr` that interfaces with the Julia `Bigsimr` package. The Julia package provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).

The subsections below describe the basic use of `bigsimr` R package by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983]. This workflow proceeds from setting up the computing environment, to data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.

@@

Here, we describe a minimal working example --- a bivariate simulation of two airquality variables: `Temperature`, in degrees Fahrenheit, and `Ozone` level, in parts per billion.

```{r ch030-air-quality-data, echo=TRUE}
df <- airquality %>% select(Temp, Ozone) %>% drop_na()
glimpse(df)
```

@@

The user must decide how to describe correlation based on the particulars of the problem. For non-normal data in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ or Kendall's $\tau$ correlation matrix $R_K$. We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for the distribution at hand (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance). Note that these estimation methods are classical approaches, not specifically designed for high-dimensional correlation estimation.

@@

Once a target correlation matrix is specified, first convert the dependency type to Pearson to correctly specify the MVN inputs. Then, check that the converted matrix $R_X$ is a valid correlation matrix (PD and 1s on the diagonal). For this bivariate example, $R_X$ is a valid correlation matrix. Note that typically in HD the resultant MVN Pearson correlation matrix is indefinite and requires approximation (see the HD examples in subsequent sections).

@@

Finally, we execute `rvec` to simulate the desired $10,000$ random vectors from the assumed bivariate distribution of `Ozone` and `Temp`. Note that the target input in specified using the pre-computed MVN correlation matrix. Figure \@ref(fig:ch030-plot-sim) plots the 10,000 simulated points.

@@

@@

\clearpage

### 2021-06-03 040-monte-carlo-evaluation.Rmd
Final push before pre-JSS submit internal review

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example. Thus, we proceed in increasing departure from normality, leading to the multivariate negative binomial (MVNB) model for our motivating data. We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for each correlation type, to evaluate the simulation's ability to simulate all possible correlations. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).

@@

<!-- 
Table: (\#tab:sims) Identical margin, bivariate simulation configurations to evaluate matching accuracy.
\begin{table}[]
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}
-->

@@

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results. Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios.

@@

Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures. Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.

@@

With information of our method's accuracy from a low-dimensional perspective, we now assess whether `bigsimr` can scale to larger dimensional problems with practical computation times. We ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, {Pearson, Spearman, Kendall} while timing the algorithm's major steps.

We first produce a synthetic "data set" by completing the following steps:

@@

```{r tab:ch040-moderateDtab}
dat_long <- benchmark_dependences %>%
  select(-total_time, -n_sim, -needed_near_pd) %>%
  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
  mutate(dim = factor(dim),
         Step = factor(Step,
                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
                       labels = c("Compute Correlation",
                                  "Adjust Correlation",
                                  "Check Admissibility",
                                  "Simulate Data"))) %>%
    rename(Correlation = corr_type, Dimensions = dim)

>>>>

We select bivariate simulation configurations to ultimately simulate our motivating discrete-valued RNA-seq example. Thus, we proceed in increasing departure from normality, leading to a multivariate negative binomial (MVNB) model for our motivating data. We begin with empirically evaluating the dependency matching across all three supported correlations --- Pearson, Spearman, and Kendall --- in identical, bivariate marginal configurations. For each pair of identical margins, we vary the target correlation across $\Omega$, the set of possible admissible values for each correlation type, to evaluate the simulation's ability to simulate all possible correlations. The simulations progress from bivariate normal, to bivariate gamma (non-normal yet continuous), and bivariate negative binomial (non-normal and discrete).

@@

\begin{table}[]
\centering
\caption{ \label{tab:sims} Identical margin, bivariate simulation configurations to evaluate matching accuracy.}
\begin{tabular}{@{}lcr@{}}
\toprule
Simulation Reps $B$ & Correlation Types & Identical-margin 2D distribution \\ \midrule
$1,000$ & Pearson ($\rho_P$) & ${\bf Y} \sim MVN( \mu= 0 , \sigma = 1, \rho_i ), i=1,\ldots,100$ \\
$10,000$ & Spearman ($\rho_S$) & ${\bf Y} \sim MVG( shape = 10, rate = 1, \rho_i ), i=1,\ldots,100$ \\
$100,000$ & Kendall ($\tau$) & ${\bf Y} \sim MVNB(p = 3\times10^{-4}, r = 4,\rho_i), i=1,\ldots,100$ \\ \bottomrule
\end{tabular}
\end{table}

@@

Figure \@ref(fig:ch040-bPlot) displays the aggregated bivariate simulation results. Table \@ref(tab:ch040-BiError) contains the mean absolute error (MAE) in reproducing the desired dependency measures for the three bivariate scenarios. Overall, the studies show that our methodology is generally accurate across the entire range of possible correlation for all three dependency measures. Our Pearson matching performs nearly as well as Spearman or Kendall, except for a slight increase in error for negative binomial case.

@@

@@

With information of our method's accuracy from a low-dimensional perspective, we now assess whether `bigsimr` can scale to larger dimensional problems with practical computation times. We ultimately generate $B=1,000$ random vectors for $d=\{100, 250, 500, 1000, 2500, 5000, 10000\}$ for each correlation type, {Pearson, Spearman, Kendall} while timing the algorithm's major steps. We begin by producing a synthetic "data set" by completing the following steps:

@@

```{r ch040-moderateDtab}
dat_long <- benchmark_dependences %>%
  select(-total_time, -n_sim, -needed_near_pd) %>%
  pivot_longer(cols = c(corr_time:sim_time), names_to = "Step", values_to = "Time") %>%
  mutate(dim = factor(dim),
         Step = factor(Step,
                       levels = c("corr_time", "adj_time", "admiss_time", "sim_time"),
                       labels = c("Compute Correlation",
                                  "Adjust Correlation",
                                  "Check Admissibility",
                                  "Simulate Data"))) %>%
    rename(Correlation = corr_type, Dimensions = dim)
